### 2020-10-13 manuscript/manuscript.rmd
update minor changes of text and figure orders

---
bibliography: manuscript.bib
csl: nature.csl
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
  word_document: default
  github_document: default
---

@@

There has been considerable debate about the placement of the root in the animal tree of life, which has emerged as one of the most difficult problems in animal phylogenetics in part because this is the oldest part of the animal tree. This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system. Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister). These analyses differed in many respects, including genes considered, species considered, molecular evolution models, and software. Here we systematically explore this question under consistent conditions by synthesizing data and results from 15 previous phylogenomic studies and performing new standardized analyses. It had previously been suggested that site heterogeneous models favor Profera-sister, but we find that this is not the case. Profera-sister is only obtained under a narrow set of conditions when the categories are unconstrained in number and include a large number of categories that apply to a small fraction of data. Our analyses help understand the sources of variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Figs. 1B, 1E), the hypothesis that the animal root marks the divergence of Porifera (sponges) from all other animals, was widely accepted though rarely tested. By contrast, there has long been uncertainty about the placement of Ctenophora (comb jellies) in the animal tree of life [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Figs. 1A, 1D). Since then, many more relevant phylogenomic studies have been published (Fig. 2), with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1, Supplementary Information section S1). The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, demonstrating for example that outgroup and model selection can have a large impact on these results (Fig. 3A). But the extensive technical variation has also made it difficult to synthesize these results to understand the underlying causes of this sensitivity. This is because methodological variance is often confounded, for example when a particular model is only available in one software package. There are a variety of things that make this a particularly challenging problem. For one, the relevant nodes are the deepest in the animal tree of life. Other factors that have been invoked include concern that Ctenophora-sister could be an artefact of long branch attraction to outgroups [@pisani2015genomic]. While some sponges have a longer root to tip length, indicating an accelarated rate of evolution in those lineages, the relatively long stem branch of Ctenophora (*e.g.* Extended Data Figs. 3, 4), together with a more typical root-to-tip distance, is instead consistent with a more recent divergence of extant ctenophores than extant poriferans [@podar2001].

@@

High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes and taxa for their matrices. As a result, the particular genes selected for analysis vary widely (Fig. 2B, Extended Data Fig. 1). Gene sampling varies in several ways, including in the fractions of single copy orthologs (*e.g.* BUSCO genes) and ribosomal protein genes in the matrix (Extended Data Fig. 2).

@@

Ingroup taxon sampling also varies widely between studies (Fig. 2B, Extended Data Fig. 1). Sampling of ingroup taxa (animals) in early studies was biased toward Bilateria. Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time as studies add additional taxa from these poorly studied groups (Extended Data Fig. 1). Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Extended Data Fig. 1). This variation is in part because newer sequencing technologies in more recent studies are usually not applied to the exact same species that were in earlier studies.

@@

Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanimalia sampling. We found that for Whelan2017 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model (Extended Data Fig. 3E). Moreover, we also found all results strongly supported Ctenophora-sister with GTR+nCAT60 models.

These results further clarify when analyses support Porifera-sister (*e.g.* Whelan2017 Extended Data Fig. 3; Philippe2009 Extended Data Fig. 4). Outgroup sampling must be restricted (Choanimalia), a Poisson rather than GTR exchange matrix must be used, and a very large number of site categories must be used (unconstrained CAT, giving hundreds of equilibrium frequency categories). Analyses under other conditions either support Ctenophora-sister or are unresolved. This is consistent across published analyses and our new panels of analyses. The question, then, isn't why similar analyses give different results, but how we should interpret variation in results when we run analyses that differ in these specific respects. If we fix the first two features to conditions that are necessary for Porifera-sister support (Choanimalia taxon sampling and a Poisson exchange matrix), there are several things we can learn from examining how the number of equilibrium frequency categories impacts results that sheds light on interpretation of those results.

There are several impacts of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister). It could be that there are a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categories are available. This would suggest that we should favor the result under a very large number of categories, Porifera-sister. Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameters. This would suggest that we should favor the result under a smaller number of categories, Ctenophora-sister. One way to assess this is to look at the allocation of sites to equilibrium frequency categories. Each new category adds the same number of parameters, but that category can be allocated to any number of sites. If as we add categories those categories are allocated to a small fraction of sites, then the cost of those additional categories is high relative to the fraction of the data they help explain, and we should be skeptical of the results when adding so many categories. That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig. 5A-B), which have `r n_categories_phil_cat` and `n_categories_whel_cat` categories. The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix. This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig. 5A-B). This suggests that support for Porifera-sister is due to adding a large number of parameters that apply to a small fraction of the data.

@@

#### Outgroup taxa sampling with C60 and CAT models
Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia, Holozoa and Opisthokonta. These datasets include the same set of genes but differ in the composition of outgroup species. Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related holozoans; Opistokonta also includes Fungi. For each Choanimalia data matrice, both C60 models in IQ-TREE and Poisson+CAT models in PhyloBayes were conducted. The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQ-TREE. Moreover, bayesian inference with the site-heterogeneous Poisson+CAT model was done with PhyloBayes-MPI v1.8. To minimize computational burden, GTR+CAT models were only performed in the representative Choanimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.

@@

For Poisson+CAT and LG in PhyloBayes, we only considered the Philippe_2009 and Whelan_2017 datasets due to computational efficiency. Since the default option in PhyloBayes does not provide the feature to calculate site-wise log likelihood for every generation, we replaced the line “int sitelogl = 0” with “int sitelogl = 1” in the file named “ReadSample.cpp” and installed PhyloBayes 4.1c so that site-wise log likelihood value can be stored to a file that ends with ".sitelogl" (via readpb –sitelogl). For each condition, we first calculated site-wise log likelihoods for each of two hypotheses (T1 and T2) using pb (via –T) and then stored site-wise log likelihood (a total number of samples for each site is 20) every ten until 300th generations, after discarding the first 100 generations using readpb (via -sitelogl -x 100 10 300). Next, we normalized site-wise log likelihood value across 20 samples for each of two hypotheses (T1 and T2) and combined normalized site-wise log likelihood values of T1 and T2 into a single file that was used to calculate gene-wise log-likelihood scores between T1 and T2 with Perl scripts from a previous study [@shen2017contentious].

#### Sensitivity analyses with different number of substitutional categories
To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60,70,80,90,110,120,150,180,360), Whelan2017_full_Choanimalia (ncat=60,340,380,420,460), Philippe2009 (ncat=60,90,120,150,180) Philippe2009_holozoa (ncat=360,400,440,480) and Ryan2013_est (ncat=60). To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Choanimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full. All PhyloBayes analyses were carried out using the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), except when a different number of categories was used.

@@

@@

We thank the Yale Center for Research Computing for use of the research computing infrastructure, specifically the Farnam HPC cluster.

@@

**Extended Data Table 1.** Table 1. Overview of data matrices used in this study.
```{r table_study_summary}
table_study_summary = read_tsv("../data_processed/tables/study_summary.tsv")
opts <- options(knitr.kable.NA = "")
knitr::kable(table_study_summary)

@@

ggsave("figures/Extended_Data_Fig1_raw.pdf", width=20, height=12)

**Extended Data Fig. 1.** Pairwise overlap between each of the primary matrices considered here. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled. The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.

@@

ggsave("figures/Extended_Data_Fig2_raw.pdf", width=10, height=4)

@@

**Extended Data Fig. 2.** Annotation and representation of BUSCO and ribosomal genes in each data matrix. (A). The number of partitions with ribosomal annotations in each matrix, relative to the number of partitions. (C). The number of partitions with annotations of BUSCO genes in each matrix, relative to the number of partitions.

ggsave("figures/Extended_Data_Fig3_raw.pdf", width=10, height=8)
```
![](figures/Extended_Data_Fig3_manuscript.png)
**Extended Data Fig. 3.** Comparison of the tree topologies under different substitutional models in Whelan2017_strict data matrix. (A). WAG+C60. (B). Poisson+nCAT60. (C). Poisson+nCAT90. (D). Poisson+CAT. (E). GTR+CAT.

tree1= read.tree("../trees_new/iqtree/Philippe2009_only_choanozoa.WAG+C60.treefile")
tree1_rooted =  root(tree1, outgroup=outgroup$relabelled_name, resolved.root=TRUE)
p1 = ggtree(tree1_rooted) +
  geom_cladelabel(node=85, label="Ctenophora") +
  geom_cladelabel(node=76, label="Porifera", angle=90, hjust='center', offset.text=.01) +
  geom_cladelabel(node=33, label="Placozoa") +
  geom_cladelabel(node=53, label="Bilateria") +
  geom_cladelabel(node=51, label="Cnidaria") +
  geom_hilight(node=85, fill="#E69F00") +
  geom_hilight(node=76, fill="#009E73") +
  geom_hilight(node=33, fill="#82D1F4") +
  geom_hilight(node=53, fill="#E6BEFF") +
  geom_hilight(node=51, fill="#911EB4") +
  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 95), hjust=-.3, size=2) + ggtitle("    WAG+C60")

tree2=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT60.con.tre")
tree2_rooted =  root(tree2, outgroup$relabelled_name, resolved.root=TRUE)
p2= ggtree(tree2_rooted) +
  geom_cladelabel(node=83, label="Ctenophora") +
  geom_cladelabel(node=85, label="Porifera", angle=90, hjust='center', offset.text=.01) +
  geom_cladelabel(node=32, label="Placozoa") +
  geom_cladelabel(node=58, label="Bilateria") +
  geom_cladelabel(node=52, label="Cnidaria") +
  geom_hilight(node=83, fill="#E69F00") +
  geom_hilight(node=85, fill="#009E73") +
  geom_hilight(node=32, fill="#82D1F4") +
  geom_hilight(node=58, fill="#E6BEFF") +
  geom_hilight(node=52, fill="#911EB4") +
  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 0.95), hjust=-.3, size=2) + ggtitle("    Poisson+nCAT60")

tree3=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT150.con.tre")
tree3_rooted =  root(tree3, outgroup$relabelled_name, resolved.root=TRUE)
p3=ggtree(tree3_rooted) +
  geom_cladelabel(node=83, label="Ctenophora") +
  geom_cladelabel(node=85, label="Porifera", angle=90, hjust='center', offset.text=.01) +
  geom_cladelabel(node=32, label="Placozoa") +
  geom_cladelabel(node=58, label="Bilateria") +
  geom_cladelabel(node=52, label="Cnidaria") +
  geom_hilight(node=83, fill="#E69F00") +
  geom_hilight(node=85, fill="#009E73") +
  geom_hilight(node=32, fill="#82D1F4") +
  geom_hilight(node=58, fill="#E6BEFF") +
  geom_hilight(node=52, fill="#911EB4") +
  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 0.95), hjust=-.3, size=2) + ggtitle("    Poisson+nCAT150")

tree4=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.bpcomp.con.tre")
tree4_rooted =  root(tree4, outgroup$relabelled_name, resolved.root=TRUE)
p4=ggtree(tree4_rooted) +
  geom_cladelabel(node=58, label="Ctenophora") +
  geom_cladelabel(node=84, label="Porifera", angle=90, hjust='center', offset.text=.01) +
  geom_cladelabel(node=35, label="Placozoa") +
  geom_cladelabel(node=60, label="Bilateria") +
  geom_cladelabel(node=52, label="Cnidaria") +
  geom_hilight(node=58, fill="#E69F00") +
  geom_hilight(node=84, fill="#009E73") +
  geom_hilight(node=35, fill="#82D1F4") +
  geom_hilight(node=60, fill="#E6BEFF") +
  geom_hilight(node=52, fill="#911EB4") +
  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 0.95), hjust=-.3, size=2) + ggtitle("    Poisson+CAT")

  tree5=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.phy_GTR_CAT.con.tre")
  tree5_rooted =  root(tree5, outgroup$relabelled_name, resolved.root=TRUE)
  p5=ggtree(tree5_rooted) +
    geom_cladelabel(node=58, label="Ctenophora") +
    geom_cladelabel(node=85, label="Porifera", angle=90, hjust='center', offset.text=.01) +
    geom_cladelabel(node=38, label="Placozoa") +
    geom_cladelabel(node=61, label="Bilateria") +
    geom_cladelabel(node=52, label="Cnidaria") +
    geom_hilight(node=58, fill="#E69F00") +
    geom_hilight(node=85, fill="#009E73") +
    geom_hilight(node=38, fill="#82D1F4") +
    geom_hilight(node=61, fill="#E6BEFF") +
    geom_hilight(node=52, fill="#911EB4") +
    geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 0.95), hjust=-.3, size=2) + ggtitle("    GTR+CAT")
  p5 = flip(p5, 85, 38)

ggsave("figures/Extended_Data_Fig4_raw.pdf", width=10, height=8)
```
![](figures/Extended_Data_Fig4_manuscript.png)
**Extended Data Fig. 4.** Comparison of the tree topologies under different substitutional models in Philippe2009_Choanimalia data matrix. (A). WAG+C60. (B). Poisson+nCAT60. (C). Poisson+nCAT150. (D). Poisson+CAT. (E). GTR+CAT.

@@

@@

![](figures/Extended_Data_Fig4_manuscript.png)

@@

![](figures/Extended_Data_Fig5.png)

@@

![](figures/Extended_Data_Fig6.png)

>>>>

---
bibliography: manuscript.bib
csl: nature.csl
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
  word_document: default
  github_document: default
---

@@

There has been considerable debate about the placement of the root in the animal tree of life, which has emerged as one of the most difficult problems in animal phylogenetics in part because this is the oldest part of the animal tree. This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system. Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister). These analyses differed in many respects, including genes considered, species considered, molecular evolution models, and software. Here we systematically explore this question under consistent conditions by synthesizing data and results from 15 previous phylogenomic studies and performing new standardized analyses. It had previously been suggested that site heterogeneous models favor Porifera-sister, but we find that this is not the case. Porifera-sister is only obtained under a narrow set of conditions when the categories are unconstrained in number and include a large number of categories that apply to a small fraction of data. Our analyses help understand the sources of variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Figs. 1B, 1E), the hypothesis that the animal root marks the divergence of Porifera (sponges) from all other animals, was widely accepted though rarely tested. By contrast, there has long been uncertainty about the placement of Ctenophora (comb jellies) in the animal tree of life [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Figs. 1A, 1D). Since then, many more relevant phylogenomic studies have been published (Fig. 2), with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1, Supplementary Information section S1). The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, demonstrating for example that outgroup and model selection can have a large impact on these results (Fig. 3A). But the extensive technical variation has also made it difficult to synthesize these results to understand the underlying causes of this sensitivity. This is because methodological variance is often confounded, for example when a particular model is only available in one software package. There are a variety of things that make this a particularly challenging problem. For one, the relevant nodes are the deepest in the animal tree of life. Other factors that have been invoked include concern that Ctenophora-sister could be an artefact of long-branch attraction to outgroups [@pisani2015genomic]. While some sponges have a longer root to tip length, indicating an accelarated rate of evolution in those lineages, the relatively long stem branch of Ctenophora (*e.g.* Extended Data Figs. 1, 2), together with a more typical root-to-tip distance, is instead consistent with a more recent divergence of extant ctenophores than extant poriferans [@podar2001].

@@

High-throughput sequencing allows investigators to readily assemble matrices with hundreds or thousands of protein-coding genes from a broad diversity of animal species (Extended Data Table 1). Studies of animal phylogeny have used a wide variety of different approaches to identifying and selecting genes and taxa for their matrices. As a result, the particular genes selected for analysis vary widely (Fig. 2B, Extended Data Fig. 1). Gene sampling varies in several ways, including in the fractions of single copy orthologs (*e.g.* BUSCO genes) and ribosomal protein genes in the matrix (Extended Data Fig. 3).

@@

Ingroup taxon sampling also varies widely between studies (Fig. 2B, Extended Data Fig. 4). Sampling of ingroup taxa (animals) in early studies was biased toward Bilateria. Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time as studies add additional taxa from these poorly studied groups (Extended Data Fig. 4). Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Extended Data Fig. 4). This variation is in part because newer sequencing technologies in more recent studies are usually not applied to the exact same species that were in earlier studies.

@@

Due to computational limitations of GTR models, we only ran GTR+CAT and GTR+nCAT60 models on representative matrices with Choanimalia sampling. We found that for Whelan2017 matrices, support shifted from Porifera-sister with Poisson+CAT model to Ctenophora-sister using GTR+CAT model (Extended Data Fig. 5E). Moreover, we also found all results strongly supported Ctenophora-sister with GTR+nCAT60 models.

These results further clarify when analyses support Porifera-sister (*e.g.* Whelan2017 Extended Data Fig. 1; Philippe2009 Extended Data Fig. 2). Outgroup sampling must be restricted (Choanimalia), a Poisson rather than GTR exchange matrix must be used, and a very large number of site categories must be used (unconstrained CAT, giving hundreds of equilibrium frequency categories). Analyses under other conditions either support Ctenophora-sister or are unresolved. This is consistent across published analyses and our new panels of analyses. The question, then, isn't why similar analyses give different results, but how we should interpret variation in results when we run analyses that differ in these specific respects. If we fix the first two features to conditions that are necessary for Porifera-sister support (Choanimalia taxon sampling and a Poisson exchange matrix), there are several things we can learn from examining how the number of equilibrium frequency categories impacts results that sheds light on interpretation of those results.

There are several impacts of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister). It could be that there are a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categories are available. This would suggest that we should favor the result under a very large number of categories, Porifera-sister. Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameters. This would suggest that we should favor the result under a smaller number of categories, Ctenophora-sister. One way to assess this is to look at the allocation of sites to equilibrium frequency categories. Each new category adds the same number of parameters, but that category can be allocated to any number of sites. If as we add categories those categories are allocated to a small fraction of sites, then the cost of those additional categories is high relative to the fraction of the data they help explain, and we should be skeptical of the results when adding so many categories. That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig. 5A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories. The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix. This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig. 5A-B). This suggests that support for Porifera-sister is due to adding a large number of parameters that apply to a small fraction of the data.

@@

#### Outgroup taxa sampling with C60 and CAT models
Because different choices of outgroups could also affect phylogenetic inference as suggested in previous analyses, we parsed the full data matrices into three different types of outgroups: Choanimalia, Holozoa and Opisthokonta. These datasets include the same set of genes but differ in the composition of outgroup species. Choanimalia only includes choanofagellates as outgroup; Holozoa also includes more distantly related Holozoans; Opistokonta also includes Fungi. For each Choanimalia data matrice, both C60 models in IQ-TREE and Poisson+CAT models in PhyloBayes were conducted. The maximum likelihood analysis was performed using the best-fit substitution model identified as above and nodal support was assessed with 1000 ultrafast bootstrap replicates using IQ-TREE. Moreover, bayesian inference with the site-heterogeneous Poisson+CAT model was done with PhyloBayes-MPI v1.8. To minimize computational burden, GTR+CAT models were only performed in the representative Choanimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full.

@@

For Poisson+CAT and LG in PhyloBayes, we only considered the Philippe2009 and Whelan2017 matrices due to computational efficiency. Since the default option in PhyloBayes does not provide the feature to calculate site-wise log likelihood for every generation, we replaced the line “int sitelogl = 0” with “int sitelogl = 1” in the file named “ReadSample.cpp” and installed PhyloBayes 4.1c so that site-wise log likelihood value can be stored to a file that ends with ".sitelogl" (via readpb –sitelogl). For each condition, we first calculated site-wise log likelihoods for each of two hypotheses (T1 and T2) using pb (via –T) and then stored site-wise log likelihood (a total number of samples for each site is 20) every ten until 300th generations, after discarding the first 100 generations using readpb (via -sitelogl -x 100 10 300). Next, we normalized site-wise log likelihood value across 20 samples for each of two hypotheses (T1 and T2) and combined normalized site-wise log likelihood values of T1 and T2 into a single file that was used to calculate gene-wise log-likelihood scores between T1 and T2 with Perl scripts from a previous study [@shen2017contentious].

#### Sensitivity analyses with different number of substitutional categories
To explore how the number of substitutional categories may affect the phylogenetic inference related to the animal phylogeny, we conducted PhyloBayes analyses with a panel of different substitutional categories in the Whelan2017_strict (ncat=60, 70, 80, 90, 110, 120, 150, 180, 360), Whelan2017_full_Choanimalia (nCAT=60, 340, 380, 420, 460), Philippe2009 (nCAT=60, 90, 120, 150, 180) Philippe2009_Holozoa (nCAT=360, 400, 440, 480) and Ryan2013_est (ncat=60). To compare the results between Poisson+CAT and GTR+CAT and minimize computational burden, GTR+CAT and GTR+CAT60 models were only performed in the representative Choanimalia matrices from Philippe2009, Ryan2013_est and Whelan2017_full. All PhyloBayes analyses were carried out using the same settings as above (see Outgroup taxa sampling with C60 and CAT models section), except when a different number of categories was used.

@@

#### Cross-validation analyses
Bayesian cross-validation implemented in PhyloBayes-MPI was used to compare the fit of Poisson+nCAT60 and Poisson+CAT models coupled with a gamma distribution of site-rate heterogeneity in Whelan2017_strict and Philippe2009_Choanimalia data matrices. Ten replicates were run, each replicate consisting of a random subsample of 10,000 sites for training the model and 2,000 sites for computing the cross-validation likelihood score. For each  run, 5,000 generations were run and the first 2,000 generations were discarded as burn-in.

@@

We thank members of the Dunn and Rokas laboratories for discussions and comments. We are grateful to Nicolas Lartillot, Kyle David and Nathan Whelan for feedback on an earlier version of this manuscript. We thank the Yale Center for Research Computing for use of the research computing infrastructure, specifically the Farnam HPC cluster. Yuanning Li was previously supported by a scholarship from the China Scholarship Council (CSC) for studying and living abroad.

@@

**Extended Data Table 1.** Table 1. Overview of data matrices used in this study.
```{r table_study_summary}
table_study_summary = read_tsv("../data_processed/tables/study_summary.tsv")
opts <- options(knitr.kable.NA = "")
knitr::kable(table_study_summary) %>%
  kable_styling(full_width = F)

@@

ggsave("figures/Extended_Data_Fig1_raw.pdf", width=10, height=8)
```
![](figures/Extended_Data_Fig1_manuscript.png)
**Extended Data Fig. 1.** Comparison of the tree topologies under different substitutional models in Whelan2017_strict data matrix. (A). WAG+C60. (B). Poisson+nCAT60. (C). Poisson+nCAT90. (D). Poisson+CAT. (E). GTR+CAT.

tree6= read.tree("../trees_new/iqtree/Philippe2009_only_choanozoa.WAG+C60.treefile")
tree6_rooted =  root(tree6, outgroup=outgroup$relabelled_name, resolved.root=TRUE)
p1 = ggtree(tree6_rooted) +
  geom_cladelabel(node=85, label="Ctenophora") +
  geom_cladelabel(node=76, label="Porifera", angle=90, hjust='center', offset.text=.01) +
  geom_cladelabel(node=33, label="Placozoa") +
  geom_cladelabel(node=53, label="Bilateria") +
  geom_cladelabel(node=51, label="Cnidaria") +
  geom_hilight(node=85, fill="#E69F00") +
  geom_hilight(node=76, fill="#009E73") +
  geom_hilight(node=33, fill="#82D1F4") +
  geom_hilight(node=53, fill="#E6BEFF") +
  geom_hilight(node=51, fill="#911EB4") +
  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 95), hjust=-.3, size=2) + ggtitle("    WAG+C60")

tree7=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT60.con.tre")
tree7_rooted =  root(tree7, outgroup$relabelled_name, resolved.root=TRUE)
p2= ggtree(tree7_rooted) +
  geom_cladelabel(node=83, label="Ctenophora") +
  geom_cladelabel(node=85, label="Porifera", angle=90, hjust='center', offset.text=.01) +
  geom_cladelabel(node=32, label="Placozoa") +
  geom_cladelabel(node=58, label="Bilateria") +
  geom_cladelabel(node=52, label="Cnidaria") +
  geom_hilight(node=83, fill="#E69F00") +
  geom_hilight(node=85, fill="#009E73") +
  geom_hilight(node=32, fill="#82D1F4") +
  geom_hilight(node=58, fill="#E6BEFF") +
  geom_hilight(node=52, fill="#911EB4") +
  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 0.95), hjust=-.3, size=2) + ggtitle("    Poisson+nCAT60")

tree8=read.tree("../trees_new/sensitive/Philippe2009_only_choanozoa.phy_Poisson_CAT150.con.tre")
tree8_rooted =  root(tree8, outgroup$relabelled_name, resolved.root=TRUE)
p3=ggtree(tree8_rooted) +
  geom_cladelabel(node=83, label="Ctenophora") +
  geom_cladelabel(node=85, label="Porifera", angle=90, hjust='center', offset.text=.01) +
  geom_cladelabel(node=32, label="Placozoa") +
  geom_cladelabel(node=58, label="Bilateria") +
  geom_cladelabel(node=52, label="Cnidaria") +
  geom_hilight(node=83, fill="#E69F00") +
  geom_hilight(node=85, fill="#009E73") +
  geom_hilight(node=32, fill="#82D1F4") +
  geom_hilight(node=58, fill="#E6BEFF") +
  geom_hilight(node=52, fill="#911EB4") +
  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 0.95), hjust=-.3, size=2) + ggtitle("    Poisson+nCAT150")

tree9=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.bpcomp.con.tre")
tree9_rooted =  root(tree9, outgroup$relabelled_name, resolved.root=TRUE)
p4=ggtree(tree9_rooted) +
  geom_cladelabel(node=58, label="Ctenophora") +
  geom_cladelabel(node=84, label="Porifera", angle=90, hjust='center', offset.text=.01) +
  geom_cladelabel(node=35, label="Placozoa") +
  geom_cladelabel(node=60, label="Bilateria") +
  geom_cladelabel(node=52, label="Cnidaria") +
  geom_hilight(node=58, fill="#E69F00") +
  geom_hilight(node=84, fill="#009E73") +
  geom_hilight(node=35, fill="#82D1F4") +
  geom_hilight(node=60, fill="#E6BEFF") +
  geom_hilight(node=52, fill="#911EB4") +
  geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 0.95), hjust=-.3, size=2) + ggtitle("    Poisson+CAT")

  tree10=read.tree("../trees_new/phylobayes/Philippe2009_only_choanozoa.phy_GTR_CAT.con.tre")
  tree10_rooted =  root(tree10, outgroup$relabelled_name, resolved.root=TRUE)
  p5=ggtree(tree10_rooted) +
    geom_cladelabel(node=58, label="Ctenophora") +
    geom_cladelabel(node=85, label="Porifera", angle=90, hjust='center', offset.text=.01) +
    geom_cladelabel(node=38, label="Placozoa") +
    geom_cladelabel(node=61, label="Bilateria") +
    geom_cladelabel(node=52, label="Cnidaria") +
    geom_hilight(node=58, fill="#E69F00") +
    geom_hilight(node=85, fill="#009E73") +
    geom_hilight(node=38, fill="#82D1F4") +
    geom_hilight(node=61, fill="#E6BEFF") +
    geom_hilight(node=52, fill="#911EB4") +
    geom_text2(aes(label=label, subset = !is.na(as.numeric(label)) & as.numeric(label) < 0.95), hjust=-.3, size=2) + ggtitle("    GTR+CAT")
  p5 = flip(p5, 85, 38)

ggsave("figures/Extended_Data_Fig2_raw.pdf", width=10, height=8)
```
![](figures/Extended_Data_Fig2_manuscript.png)
**Extended Data Fig. 2.** Comparison of the tree topologies under different substitutional models in Philippe2009_Choanimalia data matrix. (A). WAG+C60. (B). Poisson+nCAT60. (C). Poisson+nCAT150. (D). Poisson+CAT. (E). GTR+CAT.

@@

ggsave("figures/Extended_Data_Fig3_raw.pdf", width=10, height=4)

@@

**Extended Data Fig. 3.** Annotation and representation of BUSCO and ribosomal genes in each data matrix. (A). The number of partitions with ribosomal annotations in each matrix, relative to the number of partitions. (C). The number of partitions with annotations of BUSCO genes in each matrix, relative to the number of partitions.

ggsave("figures/Extended_Data_Fig4_raw.pdf", width=20, height=12)

**Extended Data Fig. 4.** Pairwise overlap between each of the primary matrices considered here. Horizontal size is proportional to the number of genes sampled, vertical size to the number of taxa sampled. The horizontal intersection shows the proportions of shared genes, the vertical intersection shows the proportions of shared taxa.

@@

ggsave("figures/Extended_Data_Fig5_raw.pdf", width=6, height=7)
```

@@

![](figures/Extended_Data_Fig6_manuscript.png)

@@

![](figures/Extended_Data_Fig7.png)

@@

![](figures/Extended_Data_Fig8.png)
